Estimating dark matter halo concentration using the integrated mass profile

### 2015-11-09 paper/paper.tex
nueva estructura para que quepa en una letter

Galaxy properties are coupled to the structural of its dark
matter halo host. 
One of the key quantities is the halo density profile which, assuming
a  Navarro-Frenk-White parameterization, can be described by the
concentration.
This has motivated the development of different algorithms to estimate
the concentraion in N-body halos.
Here we present a new algorithm to estimate the concentration
based on the radially integrated mass profile
The algorithm is easy to implement. It main advantages is that uses the full
particle information without any binning, making it realiable in
cases when low numerical resolution becomes a limitation for other
methods. 
We show the good performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods perform worst at the low resolution.
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods 
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
This dependence has motivated the detailed study of dark matter halo
properties defining its formation history and internal structure. 
In particular, the DM distribution inside a halo is key to
understand the sizes and dynamics of galaxies.

The internal matter distribution is parameterized through the density
profile. 
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
One the most popular parameterizations for DM halo radial
density distribution is the Navarro-Frenk-White (NFW) profile
\citep{NFW}.   
Furthermore, if one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution is still an
unknown), one can assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The NFW profile parameterization thus provides the conceptual
framework to study simulated dark matter halos as a function of
redshift and cosmological parameters.
Here the central quantity is the concentration. 
In these studies 
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014} 
the results are summarized through the
mass-concentration relationship; that is the distribution of
concentration values at a fixed halo mass and redshift.
The biggest challenge in such numerical experiments is having a
reliable algorithm to estimate the concentration. 
Such an algorithm should provide unbiased results and must be robust
when applied to simulations of different numerical resolution.

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This results in a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters, providing a new independent
method to estimate the concentration parameter. 
Here we provide a detailed description of this method and a
comparison against the two traditional methods by measuring its
performance both on mock halos and N-body halos.

@@

We now review the key assumptions of the different methods used
to estimate the concentration.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a function of the radial coordinate.
This method has been used broadly for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

Our method uses the integrated mass profile.  
First we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. In
this process we discard the particle at the center.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.70\textwidth]{error.pdf}
\end{center}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. Our method provide the most accurate estimate at fixed
  particle number $N$.  {\it Con respecto a esta figura tengo una
    pregunta. Yo ajusté los perfiles con menos de 200 partulas? no
    recuerdo!}
    \label{fig:error}}
\end{figure*}

@@

@@

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.33\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.33\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.33\textwidth]{conc_dens_vel.pdf}
  \end{center}
  \caption{Comparison between the concentrations measured by our
    method and the maximum velocity (left) and density (middle)
    methods. The line indicates the equal value between the two
    techniques. The right panel compares the results of the maximum
    velocity and density methods.
  \label{fig:mdv}}
\end{figure*}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.75\textwidth]{concentration.pdf}
\end{center}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  The lines correspond to the median concentration values in each bin.
  \label{fig:concentration}}
\end{figure*}

{\em Aqui hay algo importante que hay que entender antes de discutir
  esta gráfica , y es la razón por la cual el metodo del perfil de
  masa da valores de concentración tan bajos a baja masa y comparables
  a los de los otros dos métodos a alta masa. Es importante entender
  la razón de esto!.  Se me ocurren unas preguntas para tratar de
  entender. \\ 1) La grafica de c-M \ref{fig:concentration} usa todos
  los halos o solo los halos relajados? La inclusión de halos no
  relajados introduce in bias en la mediana en Cvir, así que vale la
  pena inclusive agregar una linea mas a la gráfica con los halos
  relajados y sin ellos. Para los otros dos métodos ya se sabe que
  hace la población de halos no relajados en la medida así que no creo
  necesario agregarlas para esos métodos. \\ Hay que explicar un poco
  mas la gráfica en el sentido de que se grafica y como se
  calculó. mean concentration?, Median concentrations? o mean in
  logscale?  Mas bins de cuanto? Mínimo cuantos halos por mass bin?,
  Halos relajados en todos los casos, o indistintos?}

>>>>

We present a new algorithm to estimate the concentration of N-body
dark matter halos based on the radially integrated mass profile
Its main advantage is that it uses the full particle information
without any binning, making it realiable in  cases when low numerical
resolution becomes a limitation for other methods.  Additionally, the
algorithm is straightforward to implement.  
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning to
fit the density profile. 
Tests on  the mock halos show that the accuracy of our method to recover input
concentrations varies with halo resolution, outperforming the other
two methods. For halos
sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy. 
For the maximum radial velocity method it is on the order of $20\%$
while density profile methods fail at the $100\%$ level.
For halos with $2\times 10^4$ particles our method achieves an
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We measure the mass-concentration relationship on N-body
data taking care of using halos sampled with at least $200$
particles. We find that at low masses $10^{12}\hMsun <
M<10^{13}\hMsun$ our method yields median concentration values lower
by a $15\%-20\%$ compared to the velocity and density methods 
At higher masses $3\times 10^{13}\hMsun < M < 2\times
10^{14}\hMsun$ the three methods give similar results.
\end{abstract}

@@

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
This dependence has motivated the detailed study of dark matter halo
properties such as its formation history and internal structure. 
In particular, the sizes and dynamics of galaxies are determined by
the DM distribution inside a halo.

The internal DM distribution is parameterized through the density
profile. 
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
One the most popular parameterizations for DM halo radial
density distribution is the Navarro-Frenk-White (NFW) profile
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution is still an
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides the conceptual
framework to study simulated dark matter halos as a function of
redshift and cosmological parameters.
In these studies 
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014} 
the results are summarized through the
mass-concentration relationship; that is the distribution of
concentration values at a fixed halo mass and redshift.
A big challenge in such numerical experiments is having a
reliable algorithm to estimate the concentration. 
Such an algorithm should provide unbiased results and must be robust
when applied to simulations of different numerical resolution.

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This results in a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters, providing a new independent
method to estimate the concentration parameter. 
Here we provide a detailed description of this method and a
comparison against the two traditional methods by measuring its
performance both on mock and N-body halos.

@@

We now review the key assumptions of the different methods used
to estimate the concentration before presenting our proposal.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations. 
The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

Our method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).
To build it from N-body data we follow the next steps.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. In
this process we discard the particle at the center.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. Our method provide the most accurate estimate at fixed
  particle number $N$.  {\it Con respecto a esta figura tengo una
    pregunta. Yo ajusté los perfiles con menos de 200 partulas? no
    recuerdo!}
    \label{fig:error}}
\end{figure}

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
  \end{center}
  \caption{Comparison between the concentrations measured by our
    method and the maximum velocity (left) and density (middle)
    methods. The line indicates the equal value between the two
    techniques. The right panel compares the results of the maximum
    velocity and density methods.
  \label{fig:mdv}}
\end{figure*}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{concentration.pdf}
\end{center}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  The lines correspond to the median concentration values in each bin.
  \label{fig:concentration}}
\end{figure}

@@

@@

%{\em Aqui hay algo importante que hay que entender antes de discutir
%  esta gráfica , y es la razón por la cual el metodo del perfil de
%  masa da valores de concentración tan bajos a baja masa y comparables
%  a los de los otros dos métodos a alta masa. Es importante entender
%  la razón de esto!.  Se me ocurren unas preguntas para tratar de
%  entender. \\ 1) La grafica de c-M \ref{fig:concentration} usa todos
%  los halos o solo los halos relajados? La inclusión de halos no
%  relajados introduce in bias en la mediana en Cvir, así que vale la
%  pena inclusive agregar una linea mas a la gráfica con los halos
%  relajados y sin ellos. Para los otros dos métodos ya se sabe que
%  hace la población de halos no relajados en la medida así que no creo
%  necesario agregarlas para esos métodos. \\ Hay que explicar un poco
%  mas la gráfica en el sentido de que se grafica y como se
%  calculó. mean concentration?, Median concentrations? o mean in
%  logscale?  Mas bins de cuanto? Mínimo cuantos halos por mass bin?,
%  Halos relajados en todos los casos, o indistintos?}
